---
title: "Explore KIR data"
author: "Maciej Migdal"
date: "14 04 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

devtools::load_all("MiDAS")
library(dplyr)
library(knitr)
library(kableExtra)
```

## some custom function
```{r custom_fun}
library(stringi)

hap2genes <- function(x, hap_dict) {
  x <- unlist(stringi::stri_split_fixed(x, "|"))
  x <- vapply(x, function(x) unlist(stri_split_fixed(x, pattern = "+")), character(2))
  apply(x, 2, function(i) colSums(hap_dict[i, ]))
}
```

## load data and prepare haplotype 2 genes copy numbers dictionary
```{r load_data}
kir <- read.table("KIR_output.txt", header = TRUE, fill = TRUE, stringsAsFactors = FALSE)
hap_set <- read.table("../kpi/input/HapSet18_v2.txt", header = TRUE, sep ="\t", stringsAsFactors = FALSE)
hap_dict <- dplyr::select(hap_set, X3DL3, X2DS2, X2DL2, X2DL3, X2DP1, X2DL1, X3DP1, X2DL4, X3DL1, X3DS1, X2DL5, X2DS3, X2DS5, X2DS4, X2DS1, X3DL2)
rownames(hap_dict) <- hap_set$haplotype
```

## test that hap2genes works correctly
```{r check hap2genes}
hap <- "1+3|16+3"
genes_hap2genes <- hap2genes(hap, hap_dict)

hap1 <- c("1", "3")
hap1_manual <- hap_dict[hap1, ]
hap1_manual <- colSums(hap1_manual)
hap2 <- c("16", "3")
hap2_manual <- hap_dict[hap2, ]
hap2_manual <- colSums(hap2_manual)
genes_manual <- cbind(hap1_manual, hap2_manual)
colnames(genes_manual) <- c("1+3", "16+3")

testthat::expect_equal(genes_hap2genes, genes_manual) # no errors
```

## check haplotypes ambigiuity at genes copy numbers level
```{r check_ambigiuity}
haps <- kir$Haplotypes
ambig <- c()
for (i in 1:length(haps)) {
  hap <- haps[i]
  if (nchar(hap) == 0) next()
  genes <- hap2genes(hap, hap_dict)
  if (ncol(genes) > 1) {
    nonambig <- all(apply(genes, 1, function(row) all(row == row[1])))
    if (! nonambig) {
      print(sprintf("Haplotypes for sample %i are ambigious", i))
      ambig <- append(ambig, i)
    }
  }
}

kable(kir[ambig, ], caption = "ambigious haplotypes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")


# Let's look closer into one of them to see it for our selfs
ambigious_haplotype <- kir[ambig[1], "Haplotypes"]
genes <- hap2genes(ambigious_haplotype, hap_dict)
kable(t(genes), caption = sprintf("Example %s haplotype genes copy numbers", ambigious_haplotype)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")


# Here with ambigious positions only
ambig_pos <- apply(genes, 1, function(row) ! all(row == row[1]))
kable(t(genes[ambig_pos, ]), caption = sprintf("Example %s haplotype genes copy numbers - ambigious genes only", ambigious_haplotype)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "200px")
```

# check consistency on simple presence/absence of gene level
```{r}
haps <- kir$Haplotypes
ambig <- c()
for (i in 1:length(haps)) {
  hap <- haps[i]
  if (nchar(hap) == 0) next()
  genes <- hap2genes(hap, hap_dict)
  # reduce to presence/absence
  genes <- ifelse(genes == 0, 0, 1)
  if (ncol(genes) > 1) {
    nonambig <- all(apply(genes, 1, function(row) all(row == row[1])))
    if (! nonambig) {
      print(sprintf("Haplotypes for sample %i are ambigious", i))
      ambig <- append(ambig, i)
    }
  }
}

# No ambigiutity found on presence / absence level!
```

# Conclusions:
1. In HapSet only first 16 variables (X3DL3, X2DS2, X2DL2, X2DL3, X2DP1, X2DL1, 
   X3DP1, X2DL4, X3DL1, X3DS1, X2DL5, X2DS3, X2DS5, X2DS4, X2DS1, X3DL2) 
   correspons to genes.
2. While for ambigiuous haplotypes we also observe ambiguity on gene numbers 
   level, yet they are consistient on presence / abscence level.
