---
title: "HLA protein alignments in MiDAS"
author: "Maciej Migdał"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{HLA protein alignments in MiDAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load_packages, message = FALSE}
library(midasHLA)
```

## Introduction

Several MiDAS analyses work on the level of amino acid positions rather than
whole alleles:

- the `hla_aa` experiment of `prepareMiDAS()` tests individual residues
  (e.g. `DRB1_11_V`) and, in omnibus tests, whole positions (e.g. `DRB1_11`);
- `getAllelesForAA()` and `summariseAAPosition()` show which alleles carry
  which residue at a given position;
- `hla_divergence` (Grantham distance) is computed over a fixed set of
  positions, e.g. the peptide binding groove (positions 2-182) or the B and F
  pockets.

All of them translate HLA alleles to amino acid sequences using the protein
alignments published by the IPD-IMGT/HLA Database [1, 2]. For results to be
interpretable, positions have to be numbered exactly as in the IPD-IMGT/HLA
nomenclature [3]: this is the numbering used to define structural features of
HLA molecules, such as the peptide binding pockets [4], and to report
associations, e.g. HLA-DRB1 positions 11, 71 and 74 in rheumatoid arthritis
[5] or HLA-B positions 67, 70, 97 and 156 in HIV-1 control [6]. Tools imputing
or testing HLA amino acid polymorphisms, such as SNP2HLA [7], HLA-TAPAS [8],
HATK [9] or HLAtools [10], use the same numbering.

This vignette describes the IPD-IMGT/HLA protein alignment format, how MiDAS
parses it and numbers positions, and how the parsing algorithm changed to keep
up with the evolving format of the alignments.

## IPD-IMGT/HLA protein alignments

### Where they come from

The IPD-IMGT/HLA Database [1, 2] publishes multiple sequence alignments of
all known alleles of each HLA gene with each release (currently every three
months). Protein alignments are named `<gene>_prot.txt` (e.g. `A_prot.txt`)
and are available from the
[IMGTHLA GitHub repository](https://github.com/ANHIG/IMGTHLA/tree/Latest/alignments),
which keeps a separate branch for each release (e.g. `3650` for release
3.65.0) and a `Latest` branch.

MiDAS ships pre-parsed alignments of the supported genes in its `extdata`
directory. The scripts used to download and pre-parse them are described in
the package README (section *Developers notes*).

### File layout

A (shortened) alignment file looks like this:

```
# file: A_prot.txt
# date: 2026-07-14
# version: IPD-IMGT/HLA 3.65.0
...
 Prot              -30                                1
                   |                                  |
 A*01:01:01:01           MAVM APRTLLLLLS GALAL..TQT WAG.SHS... ....MRYFFT SVSRPGRGEP ...
 A*01:01:01:02N          ---- ---------- -----..--- ---.---... ....------ ---------- ...
 A*01:01:01:03           ---- ---------- -----..--- ---.---... ....------ ---------- ...
 ...

 Prot              61
                   |
 A*01:01:01:01     DQETRNMKAH SQTDRANLGT LRGYYNQSED ...
 ...
```

- Lines starting with `#` hold metadata, including the release version.
- The alignment is split into **blocks**. Each block starts with a `Prot`
  line giving position numbers, followed by a line with `|` markers placed
  above the columns these numbers refer to. Each block then lists all alleles
  with their sequences, in groups of ten residues.
- The **first allele** in the file is the **reference allele** (e.g.
  `A*01:01:01:01`, `B*07:02:01:01`, `C*01:02:01:01`, `DRB1*01:01:01:01`). Its
  full sequence is shown; other alleles are shown relative to it.

Characters used in the sequences:

| Character | Meaning |
|---|---|
| `A`-`Z` | amino acid residue (`X` marks a stop codon) |
| `-` | same residue as in the reference allele |
| `.` | no residue in this allele at this column (a gap) |
| `*` | residue unknown, the allele was not sequenced in this region |

The IPD-IMGT/HLA REST API (`https://www.ebi.ac.uk/cgi-bin/ipd/api/allele/`)
provides the same information for individual alleles: allele's protein
sequence, signal peptide and mature protein boundaries, and a "protein CIGAR"
string describing where allele's residues are placed in the alignment (`M` -
residue, `D` - no residue, `N` - not sequenced). We use it to validate the
parser (see [Validation](#validation)).

### Position numbering

Positions are numbered according to the **reference allele** [3]:

- position 1 is the first residue of the mature protein, it is marked by the
  `1` above the first block;
- residues of the signal peptide (leader) have negative numbers: -1 is the
  residue directly preceding position 1. There is no position 0;
- every residue of the reference allele gets the next number.

Because the numbering follows the residues of the reference allele, **an
alignment column where the reference allele has a gap (`.`) has no number**.
Such a column exists only because some other allele has an extra residue
there - an **insertion** relative to the reference. In the example above
`A*01:01:01:01` reads `WAG.SHS... ....MRYFFT`: `G` is position 1, it is
followed by one insertion column, `S`, `H` and `S` are positions 2-4, followed
by seven insertion columns, and `M` is position 5.

In contrast, a gap in any other allele, in a column where the reference allele
has a residue, is a **deletion** in that allele; the column keeps its number.

A few more details matter when parsing:

- Some alleles extend beyond the last residue of the reference allele. Their
  extra residues are shown in additional columns, or in additional blocks
  listing only these alleles.
- The `Prot` lines are labels for readers rather than a strict index: a block
  starting with insertion columns is labelled with the number of the next
  numbered position, and in some files (e.g. TAP1) block labels are off by
  one. MiDAS therefore uses only the first marked position (`1`) as an anchor
  and counts the reference residues from there.

### Naming insertion positions

Insertion columns still carry information: for example all HLA-C\*17 alleles
have six extra residues (`AVLAVL`) in the transmembrane region, after position
300 of the reference allele. They need a name that does not consume a position
number - otherwise every following position would be shifted. MiDAS names
insertion columns after the preceding numbered position followed by the index
of the insertion column, e.g. `300.1`, `300.2`, ..., `300.6`. The same
convention is used by HLAtools [10]. Other tools represent insertions
differently, e.g. SNP2HLA and HLA-TAPAS encode a whole insertion as a single
marker spanning two positions, such as `INS_C_295x296_VLAVLA` [7, 8].

Note that `300.10` and `300.1` are different positions: position names must
always be treated as text, not numbers.

## Parsing algorithm

`readHlaAlignments()` parses an alignment file in the following steps:

1. Read lines with allele sequences, recognized by a valid allele name in the
   first field, and concatenate each allele's sequence over all blocks.
2. Make all sequences as long as the longest allele, so that alleles extending
   beyond the reference allele are kept. Columns beyond the end of an allele's
   sequence are marked with `""`.
3. Replace `-` with the residue of the reference allele in that column.
4. Find the alignment column of position 1, using the `1` marker in the first
   `Prot` line.
5. Number the columns: count residues of the reference allele, with position 1
   at the marked column, negative positions before it and no position 0.
   Columns where the reference allele has no residue are named after the
   preceding position: `<position>.<k>`, where `k` is the index of the
   insertion column.
6. Optionally (`trim = TRUE`) drop the signal peptide, i.e. columns before
   position 1.

The example below uses a small alignment in the IPD-IMGT/HLA format. The
reference allele `X*01:01:01:01` has gaps in four places: one in the signal
peptide, one after position 1, two after position 4 and one after position 11.

```{r toy_alignment}
toy_alignment <- c(
  "# file: X_prot.txt",
  "                   ",
  " Prot              -3   1",
  "                    |   |",
  " X*01:01:01:01      MK.AG.SHS..MRYFF ",
  " X*01:02            --.--K---..----- ",
  " X*02:01            --.--.---RR----Y ",
  " X*03:01            --.--.-.-..----- ",
  " X*04:01            --Q--.---..----- ",
  "",
  " Prot               10",
  "                    |",
  " X*01:01:01:01      TS.VS ",
  " X*01:02            --.-- ",
  " X*02:01            --.-- ",
  " X*03:01            --.-- ",
  " X*04:01            --.-. "
)
file <- tempfile(fileext = "_prot.txt")
writeLines(toy_alignment, file)

aln <- readHlaAlignments(file)
aln
```

In the resulting matrix:

- `X*01:02` has an insertion of `K` after position 1 (column `1.1`);
- `X*02:01` has an insertion of `RR` after position 4 (`4.1`, `4.2`) and a
  substitution `F9Y`;
- `X*03:01` has a deletion of position 3;
- `X*04:01` has an insertion of `Q` in the signal peptide (`-2.1`) and a
  deletion of position 13.

The columns corresponding to the reference residues are numbered exactly as
in the IPD-IMGT/HLA nomenclature, regardless of insertions in other alleles:

```{r numbered_positions}
aln[, ! grepl(".", colnames(aln), fixed = TRUE)]
```

With `trim = TRUE` the signal peptide is removed:

```{r trim}
readHlaAlignments(file, trim = TRUE)
```

### Use of positions in MiDAS

- **`hla_aa` experiment.** Insertion columns are treated like any other
  position: variables are named `<gene>_<position>_<residue>`, e.g.
  `C_300.1_A` for the first inserted residue of HLA-C\*17 alleles, and
  `C_300.1_.` for alleles without the insertion. As only variable positions
  are used, insertion columns appear in the analysis only when alleles with the
  insertion are present in the data. Omnibus tests group variables by
  position, e.g. `C_300.1`.
- **`getAllelesForAA()` and `summariseAAPosition()`** accept insertion
  positions, e.g. `getAllelesForAA(midas, "C_300.1")`.
- **Grantham distance** (`hla_divergence`) [11, 12] is computed only over
  numbered positions (e.g. 2-182 for the binding groove). Positions are
  selected by name, so insertion columns do not shift the selection; Grantham
  distance is not defined between a residue and a gap, thus insertion columns
  are not used.

For example, with alignments from IPD-IMGT/HLA release 3.65.0:

```{r insertion_example, eval = FALSE}
aln <- readHlaAlignments(gene = "C")
aln[c("C*01:02:01:01", "C*17:03:01:01"), c("300", paste0("300.", 1:6), "301")]
#>               300 300.1 300.2 300.3 300.4 300.5 300.6 301
#> C*01:02:01:01 "L" "."   "."   "."   "."   "."   "."   "G"
#> C*17:03:01:01 "L" "A"   "V"   "L"   "A"   "V"   "L"   "G"
```

## How the algorithm changed

### Original implementation

The original parser (used up to midasHLA 1.11) followed steps 1, 3, 4 and 6
of the algorithm above, but in step 5 it **numbered every alignment column**:
columns before the `1` marker received negative numbers and all following
columns received consecutive numbers. Sequences were also truncated to the
length of the reference allele.

This is correct only if the reference allele has a residue in every column,
which held for most genes in the releases available when MiDAS was developed.
Each insertion column, however, shifted the numbers of all the following
positions by one. In the shipped alignments (IPD-IMGT/HLA release 3.44.0)
this affected the C-terminal parts of a few genes, whose reference alleles
already contained insertion columns:

- HLA-B: one insertion column after position 294, so positions after it were
  shifted by one (the last position was 339 instead of 338);
- HLA-C: six insertion columns after position 300 (the HLA-C\*17 insertion),
  so positions after it were shifted by six (the last position was 348
  instead of 342):

| Reference `C*01:02:01:01` | ... | 300 | | | | | | | 301 | 302 | ... |
|---|---|---|---|---|---|---|---|---|---|---|---|
| IPD-IMGT/HLA numbering | ... | 300 | *300.1* | *300.2* | *300.3* | *300.4* | *300.5* | *300.6* | 301 | 302 | ... |
| original parser | ... | 300 | 301 | 302 | 303 | 304 | 305 | 306 | 307 | 308 | ... |

- HLA-DQB1 and MICA: similar shifts towards the C-terminus.

Positions in the peptide binding groove, the B and F pockets and therefore the
Grantham distances were not affected, which may explain why the problem went
unnoticed.

### Evolving IPD-IMGT/HLA alignments

With new alleles, the IPD-IMGT/HLA alignments gained insertion columns in
many more places, including the signal peptide and the peptide binding groove
- e.g. HLA-B has 131 insertion columns in release 3.65.0, the first ones in
the signal peptide and after position 24. When the alignments were updated to
release 3.55.0 in January 2024, the original parser shifted almost all
positions, which broke amino acid level analyses and Grantham distances. The
update was reverted until the parser could be fixed.

The `Grantham distance` calculation had an additional hidden assumption: it
selected positions 2-182 by **column index** rather than by position name,
which is equivalent only when every column is a numbered position.

### Current implementation

The parser was changed in two steps:

1. Columns where the reference allele has a gap are recognized as insertion
   columns and excluded from numbering, restoring the IPD-IMGT/HLA numbering
   of all positions.
2. Instead of being dropped, insertion columns are kept and named
   `<position>.<k>`, and the alignment spans the longest allele, so that the
   sequences of all alleles are represented completely.

Along with it:

- Grantham distance selects positions by name and uses numbered positions
  only;
- position names in `getAllelesForAA()`, `summariseAAPosition()` and
  `runMiDAS()` results accept insertion positions;
- the alignment pre-parsing script reuses the numbering of
  `readHlaAlignments()` instead of a separate copy of the original logic,
  and the download script fetches alignments of a pinned IPD-IMGT/HLA release
  from the IMGTHLA GitHub repository.

## Validation {#validation}

The parser is validated against data independent of the alignment text files:
records of the IPD-IMGT/HLA REST API, which provide each allele's protein
sequence, signal peptide length and placement of its residues in the
alignment. Package tests use excerpts of IPD-IMGT/HLA 3.65.0 alignments of
HLA-A, -B, -C and -DQB1, covering reference alleles, insertions in the signal
peptide and the transmembrane region, an allele extending far beyond the
reference allele, deletions and a partially sequenced allele, and check that:

- every column name and every residue matches the alignment built from the API
  records, numbered as described above;
- reading an allele's residues in order gives exactly its protein sequence,
  i.e. nothing is lost or reordered;
- position 1 is the first residue of the allele's mature protein.

Additionally, for all genes shipped with MiDAS we checked, using release
3.65.0, that numbered positions agree with the position labels of the `Prot`
lines and that, outside of the regions described in
[Original implementation](#original-implementation), they agree with the
previously shipped alignments (apart from a few dozen residues corrected by
IPD-IMGT/HLA between the releases). For HLA-A, -B and -C, the conserved
cysteines of class I heavy chains are found at positions 101, 164, 203 and
259. For HLA-A, -B, -C and -DQB1, numbered positions and residues also agree
with alignments built by HLAtools [10].

## References

1. Robinson J, Barker DJ, Georgiou X, Cooper MA, Flicek P, Marsh SGE.
   IPD-IMGT/HLA Database. *Nucleic Acids Research*. 2020;48(D1):D948-D955.
   [doi:10.1093/nar/gkz950](https://doi.org/10.1093/nar/gkz950)
2. Barker DJ, Maccari G, Georgiou X, Cooper MA, Flicek P, Robinson J, Marsh
   SGE. The IPD-IMGT/HLA Database. *Nucleic Acids Research*.
   2023;51(D1):D1053-D1060.
   [doi:10.1093/nar/gkac1011](https://doi.org/10.1093/nar/gkac1011)
3. Marsh SGE, Albert ED, Bodmer WF, et al. Nomenclature for factors of the
   HLA system, 2010. *Tissue Antigens*. 2010;75(4):291-455.
   [doi:10.1111/j.1399-0039.2010.01466.x](https://doi.org/10.1111/j.1399-0039.2010.01466.x)
4. Saper MA, Bjorkman PJ, Wiley DC. Refined structure of the human
   histocompatibility antigen HLA-A2 at 2.6 Å resolution. *Journal of
   Molecular Biology*. 1991;219(2):277-319.
   [doi:10.1016/0022-2836(91)90567-P](https://doi.org/10.1016/0022-2836(91)90567-P)
5. Raychaudhuri S, Sandor C, Stahl EA, et al. Five amino acids in three HLA
   proteins explain most of the association between MHC and seropositive
   rheumatoid arthritis. *Nature Genetics*. 2012;44(3):291-296.
   [doi:10.1038/ng.1076](https://doi.org/10.1038/ng.1076)
6. The International HIV Controllers Study. The major genetic determinants of
   HIV-1 control affect HLA class I peptide presentation. *Science*.
   2010;330(6010):1551-1557.
   [doi:10.1126/science.1195271](https://doi.org/10.1126/science.1195271)
7. Jia X, Han B, Onengut-Gumuscu S, et al. Imputing amino acid polymorphisms
   in human leukocyte antigens. *PLoS ONE*. 2013;8(6):e64683.
   [doi:10.1371/journal.pone.0064683](https://doi.org/10.1371/journal.pone.0064683)
8. Luo Y, Kanai M, Choi W, et al. A high-resolution HLA reference panel
   capturing global population diversity enables multi-ancestry fine-mapping
   in HIV host response. *Nature Genetics*. 2021;53(10):1504-1516.
   [doi:10.1038/s41588-021-00935-7](https://doi.org/10.1038/s41588-021-00935-7)
9. Choi W, Luo Y, Raychaudhuri S, Han B. HATK: HLA analysis toolkit.
   *Bioinformatics*. 2021;37(3):416-418.
   [doi:10.1093/bioinformatics/btaa684](https://doi.org/10.1093/bioinformatics/btaa684)
10. Tran L, Nickens R, Luu V, Petersdorf EW, Mack SJ. HLAtools, Searching Shared HLA Amino Acid Residue
    Prevalence, and the Global Frequency Browsers: New Computational Resources
    for Working With HLA Data and Visualizing Global Patterns of HLA Variation.
    *International Journal of Immunogenetics*. 2025;52(6):358-370.
    [doi:10.1111/iji.70013](https://doi.org/10.1111/iji.70013);
    R package: [CRAN](https://cran.r-project.org/package=HLAtools)
11. Grantham R. Amino acid difference formula to help explain protein
    evolution. *Science*. 1974;185(4154):862-864.
    [doi:10.1126/science.185.4154.862](https://doi.org/10.1126/science.185.4154.862)
12. Pierini F, Lenz TL. Divergent allele advantage at human MHC genes:
    signatures of past and ongoing selection. *Molecular Biology and
    Evolution*. 2018;35(9):2145-2158.
    [doi:10.1093/molbev/msy116](https://doi.org/10.1093/molbev/msy116)
13. Migdal M, Ruan DF, Forrest WF, Horowitz A, Hammer C. MiDAS—Meaningful
    Immunogenetic Data at Scale. *PLoS Computational Biology*.
    2021;17(7):e1009131.
    [doi:10.1371/journal.pcbi.1009131](https://doi.org/10.1371/journal.pcbi.1009131)
